---
title: "Similarity-based vs _derep Function Comparison"
subtitle: "Numerical verification that new .sim slot math produces equivalent results to original .dissim math"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 6,
  fig.height = 5
)
```

## Background

This notebook verifies that the new similarity-based functions (`.sim` slots,
`GetSimMat()`) produce numerically equivalent results to the original `_derep`
functions (`.dissim` slots, `GetDistanceMat()`), using the Tutorial 1
Arabidopsis demo dataset.

**Mathematical basis:** For normalized similarity $S \in [0,1]$, dissimilarity
$D = 1 - S$. The new functions rewrite all matrix operations in terms of $S$
directly, avoiding `as.matrix()` densification on large sparse matrices. The
identities used are:

- $p'Dp = 1 - p'Sp$ (Rao Q, $p$ a probability vector)
- $Dp = 1 - Sp$ (when $\sum p = 1$)
- $DP_q = \text{colSums}(P_q) - S P_q$ (general vector $P_q$)
- CSS (CSCS) $= S$ directly (was $1 - D$)

**Functions compared:**

| `_derep` | new | math changes |
|---|---|---|
| `GetFunctionalHillNumber_derep` | `GetFunctionalHillNumber` | Rao Q, Dp, DPq reformulated |
| `GetFaithPD_derep` | `GetFaithPD` | S→D for hclust; fastcluster C backend |
| `GetAlphaDiversity_derep` | `GetAlphaDiversity` | delegates to above |
| `GetBetaDiversity_derep` | `GetBetaDiversity` | CSCS = S directly; Gen.Uni S→D |
| `GenerateHeatmapInputs_derep` | `GenerateHeatmapInputs` | `as.dist(1-S)` for clustering |

> **Tree-based functions note:** `GetFaithPD` and `GetBetaDiversity(Gen.Uni)`
> both use **average-linkage** hclust on this small demo dataset (`use_mst =
> FALSE`, n < 10,000). Since $D = 1 - S$ the trees are identical, so outputs
> should match. If `use_mst = TRUE` (single-linkage for large n), results will
> intentionally differ.

---

## Setup

```{r load-pkg}
# Load from local source (run from package root) or installed package
if (requireNamespace("devtools", quietly = TRUE) && file.exists("DESCRIPTION")) {
  devtools::load_all(quiet = TRUE)
} else {
  library(ecomet)
}
library(ggplot2)
library(dplyr)
```

```{r build-mmo}
data_dir <- system.file("extdata/tutorials/treatment_based", package = "ecomet")
stopifnot(nzchar(data_dir))

mmo <- GetMZmineFeature(
  mzmine_dir   = file.path(data_dir, "feature_table_demo.csv"),
  metadata_dir = file.path(data_dir, "metadata_demo.csv"),
  group_col    = "group",
  sample_col   = "sample"
)
mmo <- LogNormalization(mmo)

demo_dreams <- file.path(data_dir, "dreams_sim_demo.csv")
mmo <- AddChemDist(mmo, dreams_dir = demo_dreams)   # -> mmo$dreams.dissim
mmo <- AddChemSim(mmo,  dreams_dir = demo_dreams)   # -> mmo$dreams.sim

cat("Features:", nrow(mmo$feature_data),
    " | Samples:", nrow(mmo$metadata), "\n")
cat("dreams.dissim range: [", round(min(mmo$dreams.dissim), 4), ",",
    round(max(mmo$dreams.dissim), 4), "]\n")
cat("dreams.sim range:    [", round(min(mmo$dreams.sim),    4), ",",
    round(max(mmo$dreams.sim),    4), "]\n")
```

```{r align-matrices}
feature  <- GetNormFeature(mmo, "Log")
metadata <- mmo$metadata
dist_mat <- GetDistanceMat(mmo, "dreams")
sim_mat  <- GetSimMat(mmo,      "dreams")

keep_ids <- intersect(feature$id, rownames(dist_mat))
feat_sub <- feature[feature$id %in% keep_ids, , drop = FALSE]
dist_sub <- dist_mat[keep_ids, keep_ids]
sim_sub  <- sim_mat[keep_ids,  keep_ids]

cat("Aligned features:", length(keep_ids), "\n")
cat("D + S = 1 check (upper triangle):",
    isTRUE(all.equal(
      as.numeric(dist_sub[upper.tri(dist_sub)] + sim_sub[upper.tri(sim_sub)]),
      rep(1, sum(upper.tri(dist_sub))), tolerance = 1e-6
    )), "\n")
```

---

## Plot helper

```{r helper}
PASS_TOL <- 1e-6

corr_plot <- function(x_vals, y_vals, label,
                      x_label = "_derep value",
                      y_label = "similarity-based value") {
  mask <- is.finite(x_vals) & is.finite(y_vals)
  x <- as.numeric(x_vals[mask])
  y <- as.numeric(y_vals[mask])

  r        <- cor(x, y, method = "pearson")
  max_diff <- max(abs(x - y))
  passed   <- max_diff < PASS_TOL

  df <- data.frame(derep = x, new = y)
  rng_x <- range(x); rng_y <- range(y)

  status_label <- if (passed) "PASS" else "FAIL"
  status_color <- if (passed) "#27ae60" else "#e74c3c"

  sub_text <- sprintf(
    "Pearson r = %.8f   |   Max |delta| = %.2e   |   n = %d points",
    r, max_diff, length(x)
  )

  ann_x <- rng_x[1] + 0.04 * diff(rng_x)
  ann_y <- rng_y[2] - 0.04 * diff(rng_y)

  p <- ggplot(df, aes(x = derep, y = new)) +
    geom_abline(slope = 1, intercept = 0,
                color = "firebrick", linetype = "dashed", linewidth = 0.7) +
    geom_point(alpha = 0.55, color = "#2980b9", size = 1.8) +
    annotate("label",
             x = ann_x, y = ann_y,
             label = status_label,
             hjust = 0, vjust = 1,
             fill  = status_color, color = "white",
             fontface = "bold", size = 4.5, label.size = NA) +
    labs(
      title    = label,
      subtitle = sub_text,
      x        = x_label,
      y        = y_label,
      caption  = "Red dashed line = perfect agreement (y = x)"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title    = element_text(face = "bold", size = 12),
      plot.subtitle = element_text(size = 8.5, color = "gray40"),
      plot.caption  = element_text(size = 8,   color = "gray55")
    )

  list(plot = p, r = r, max_diff = max_diff, passed = passed,
       label = label)
}

# Accumulate results for final summary
results <- list()
show_result <- function(res) {
  results[[res$label]] <<- res
  print(res$plot)
  cat(sprintf("  r = %.10f  |  max|delta| = %.3e  |  %s\n\n",
              res$r, res$max_diff, if (res$passed) "PASS" else "FAIL"))
}
```

---

## 1 — GetFunctionalHillNumber

**What changed:** Rao Q rewritten as $1 - p'Sp$ (avoids $D = 1 - S$ densification).
For $q = 1$: $Dp = 1 - Sp$. For $q \neq 1$: $DP_q = \text{colSums}(P_q) - SP_q$.

```{r fn-hill-q1}
h_d1 <- GetFunctionalHillNumber_derep(feat_sub, metadata, dist_sub, q = 1)
h_n1 <- GetFunctionalHillNumber(      feat_sub, metadata, sim_sub,  q = 1)

show_result(corr_plot(h_d1$value, h_n1$value,
  label   = "GetFunctionalHillNumber  q = 1",
  x_label = "_derep: distance-based value",
  y_label = "new: similarity-based value"))
```

```{r fn-hill-q2}
h_d2 <- GetFunctionalHillNumber_derep(feat_sub, metadata, dist_sub, q = 2)
h_n2 <- GetFunctionalHillNumber(      feat_sub, metadata, sim_sub,  q = 2)

show_result(corr_plot(h_d2$value, h_n2$value,
  label   = "GetFunctionalHillNumber  q = 2",
  x_label = "_derep: distance-based value",
  y_label = "new: similarity-based value"))
```

---

## 2 — GetFaithPD

**What changed:** similarity converted to distance ($1 - S$) inside the function
before `hclust`. For $n \leq 10{,}000$ with `use_mst = FALSE` both versions use
average-linkage on the same underlying distances, so trees and PD values are
identical.

```{r faith-pd}
fpd_d <- GetFaithPD_derep(feat_sub, metadata, dist_sub)
fpd_n <- GetFaithPD(      feat_sub, metadata, sim_sub, use_mst = FALSE)

show_result(corr_plot(fpd_d$PD, fpd_n$PD,
  label   = "GetFaithPD  (PD, average linkage)",
  x_label = "_derep: distance-based PD",
  y_label = "new: similarity-based PD"))
```

### FaithPD: MST vs average linkage (informational — values intentionally differ)

```{r faith-pd-mst, fig.width=6, fig.height=4.5}
fpd_mst <- suppressMessages(
  GetFaithPD(feat_sub, metadata, sim_sub, use_mst = TRUE)
)

# Not a PASS/FAIL check — just show how much the tree choice matters
df_mst <- data.frame(
  avg_linkage   = fpd_d$PD,
  single_linkage = fpd_mst$PD,
  sample         = fpd_d$sample
)

ggplot(df_mst, aes(x = avg_linkage, y = single_linkage, label = sample)) +
  geom_abline(slope = 1, intercept = 0,
              color = "firebrick", linetype = "dashed", linewidth = 0.7) +
  geom_point(color = "#8e44ad", size = 2.5, alpha = 0.8) +
  labs(
    title    = "FaithPD: average linkage vs MST/single-linkage (informational)",
    subtitle = sprintf("Pearson r = %.4f  |  These SHOULD differ — different tree topologies",
                       cor(df_mst$avg_linkage, df_mst$single_linkage)),
    x = "Average-linkage PD (_derep)",
    y = "MST/single-linkage PD (use_mst = TRUE)",
    caption = "MST path used automatically when n > 10,000. Shown here on small data for illustration."
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 8.5, color = "gray40"))
```

---

## 3 — GetAlphaDiversity

**What changed:** uses `GetSimMat()` instead of `GetDistanceMat()`; delegates to
the updated `GetFunctionalHillNumber` and `GetFaithPD`.

```{r alpha-weighted-q1}
a_d <- GetAlphaDiversity_derep(mmo, mode = "weighted", q = 1,
                               normalization = "Log", distance = "dreams",
                               output = "sample_level")
a_n <- GetAlphaDiversity(mmo,       mode = "weighted", q = 1,
                         normalization = "Log", distance = "dreams",
                         output = "sample_level", use_mst = FALSE)

show_result(corr_plot(a_d$value, a_n$value,
  label   = "GetAlphaDiversity  mode = weighted, q = 1",
  x_label = "_derep value",
  y_label = "new value"))
```

```{r alpha-weighted-q2}
a_d2 <- GetAlphaDiversity_derep(mmo, mode = "weighted", q = 2,
                                normalization = "Log", distance = "dreams",
                                output = "sample_level")
a_n2 <- GetAlphaDiversity(mmo,        mode = "weighted", q = 2,
                          normalization = "Log", distance = "dreams",
                          output = "sample_level", use_mst = FALSE)

show_result(corr_plot(a_d2$value, a_n2$value,
  label   = "GetAlphaDiversity  mode = weighted, q = 2",
  x_label = "_derep value",
  y_label = "new value"))
```

```{r alpha-faith}
af_d <- GetAlphaDiversity_derep(mmo, mode = "faith",
                                normalization = "Log", distance = "dreams",
                                output = "sample_level")
af_n <- GetAlphaDiversity(mmo,        mode = "faith",
                          normalization = "Log", distance = "dreams",
                          output = "sample_level", use_mst = FALSE)

show_result(corr_plot(af_d$value, af_n$value,
  label   = "GetAlphaDiversity  mode = faith",
  x_label = "_derep value",
  y_label = "new value"))
```

---

## 4 — GetBetaDiversity: CSCS

**What changed:** CSCS used `CSS = 1 - scaled_dissimilarity`; new version uses
`CSS = scaled_similarity` directly (algebraically identical since $S = 1 - D$).

```{r beta-cscs}
bc_d <- GetBetaDiversity_derep(mmo, method = "CSCS",
                               normalization = "Log", distance = "dreams")
bc_n <- GetBetaDiversity(mmo,       method = "CSCS",
                         normalization = "Log", distance = "dreams")

ut <- upper.tri(bc_d)
show_result(corr_plot(bc_d[ut], bc_n[ut],
  label   = "GetBetaDiversity  method = CSCS  (upper triangle pairwise)",
  x_label = "_derep pairwise dissimilarity",
  y_label = "new pairwise dissimilarity"))
```

---

## 5 — GetBetaDiversity: Generalized UniFrac

**What changed:** same S→D conversion as FaithPD; both use average-linkage at
small n, so the compound tree is identical and all three weighting levels match.

```{r beta-genuni}
bg_d <- suppressMessages(
  GetBetaDiversity_derep(mmo, method = "Gen.Uni",
                         normalization = "Log", distance = "dreams")
)
bg_n <- suppressMessages(
  GetBetaDiversity(mmo,       method = "Gen.Uni",
                   normalization = "Log", distance = "dreams", use_mst = FALSE)
)

for (nm in c("d_0", "d_0.5", "d_1")) {
  ut <- upper.tri(bg_d[[nm]])
  show_result(corr_plot(bg_d[[nm]][ut], bg_n[[nm]][ut],
    label   = sprintf("GetBetaDiversity  Gen.Uni  %s  (upper triangle)", nm),
    x_label = "_derep pairwise distance",
    y_label = "new pairwise distance"))
}
```

---

## 6 — GetBetaDiversity: Bray-Curtis & Jaccard

**What changed:** nothing — these methods don't use a chemical similarity
matrix. Outputs should be **bit-for-bit identical**.

```{r beta-bray}
bb_d <- GetBetaDiversity_derep(mmo, method = "bray",    normalization = "Log")
bb_n <- GetBetaDiversity(mmo,       method = "bray",    normalization = "Log")
ut   <- upper.tri(bb_d)

show_result(corr_plot(bb_d[ut], bb_n[ut],
  label   = "GetBetaDiversity  method = bray  (no matrix used — must be identical)",
  x_label = "_derep Bray-Curtis",
  y_label = "new Bray-Curtis"))
```

```{r beta-jaccard}
bj_d <- GetBetaDiversity_derep(mmo, method = "jaccard", normalization = "PA")
bj_n <- GetBetaDiversity(mmo,       method = "jaccard", normalization = "PA")
ut   <- upper.tri(bj_d)

show_result(corr_plot(bj_d[ut], bj_n[ut],
  label   = "GetBetaDiversity  method = jaccard  (no matrix used — must be identical)",
  x_label = "_derep Jaccard",
  y_label = "new Jaccard"))
```

---

## 7 — GenerateHeatmapInputs

**What changed:** uses `GetSimMat()` and builds `as.dist(1 - sim_matrix)` for
row clustering. `FC_matrix` (the abundance data) is unchanged; `dist_matrix`
(the clustering distances) should equal the _derep version since $D = 1 - S$.

```{r heatmap-inputs}
hm_d <- GenerateHeatmapInputs_derep(mmo, normalization = "Log", distance = "dreams")
hm_n <- GenerateHeatmapInputs(mmo,       normalization = "Log", distance = "dreams")

# FC_matrix: feature x group abundance data — no matrix involved, must be identical
show_result(corr_plot(
  as.numeric(hm_d$FC_matrix), as.numeric(hm_n$FC_matrix),
  label   = "GenerateHeatmapInputs  FC_matrix (group means, no chemical matrix)",
  x_label = "_derep group mean",
  y_label = "new group mean"))
```

```{r heatmap-dist}
# dist_matrix: distances used for row clustering (derived from similarity/dissimilarity)
show_result(corr_plot(
  as.numeric(hm_d$dist_matrix), as.numeric(hm_n$dist_matrix),
  label   = "GenerateHeatmapInputs  dist_matrix (clustering distances, D = 1 - S)",
  x_label = "_derep clustering distance",
  y_label = "new clustering distance"))
```

---

## Summary

```{r summary, fig.width=8, fig.height=max(3, length(results) * 0.5 + 1)}
summary_df <- data.frame(
  comparison = names(results),
  r          = sapply(results, `[[`, "r"),
  max_diff   = sapply(results, `[[`, "max_diff"),
  passed     = sapply(results, `[[`, "passed"),
  stringsAsFactors = FALSE
)
summary_df$status <- ifelse(summary_df$passed, "PASS", "FAIL")

# Print table
cat(sprintf("\n%-70s  %12s  %10s  %6s\n",
            "Comparison", "Pearson r", "Max |delta|", "Status"))
cat(strrep("-", 106), "\n")
for (i in seq_len(nrow(summary_df))) {
  cat(sprintf("%-70s  %12.8f  %10.2e  %6s\n",
              substr(summary_df$comparison[i], 1, 70),
              summary_df$r[i],
              summary_df$max_diff[i],
              summary_df$status[i]))
}
cat(strrep("-", 106), "\n")
cat(sprintf("PASS: %d / %d\n", sum(summary_df$passed), nrow(summary_df)))

# Bar chart of max |delta| per comparison
ggplot(summary_df, aes(
    x    = reorder(comparison, max_diff),
    y    = max_diff,
    fill = status)) +
  geom_col() +
  geom_hline(yintercept = PASS_TOL, linetype = "dashed",
             color = "gray30", linewidth = 0.7) +
  annotate("text", x = 0.6, y = PASS_TOL * 1.5,
           label = paste("tolerance =", PASS_TOL),
           hjust = 0, size = 3.2, color = "gray30") +
  scale_fill_manual(values = c("PASS" = "#27ae60", "FAIL" = "#e74c3c")) +
  scale_y_log10() +
  coord_flip() +
  labs(
    title = "Max absolute difference: _derep vs similarity-based",
    subtitle = "Dashed line = pass/fail threshold (1e-6). Log scale.",
    x = NULL, y = "Max |delta|  (log scale)", fill = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title    = element_text(face = "bold"),
    legend.position = "top"
  )
```
